---
title: "Data Exploration"
author: "Emam Hossain"
date: "`r Sys.Date()`"
output:
  html_notebook: default
  pdf_document: default
---

# Univariate Exploration

# Load the Iris Dataset
```{r}
# Load dataset
data(iris)

# View first few rows
head(iris)

```

# Summary Statistics
```{r}
# Summary of numeric variables
summary(iris)

```
# Univariate Plots
```{r}
# Histogram (Distribution of a Numeric Variable)
# Histogram of Sepal Length
hist(iris$Sepal.Length, 
     main="Histogram of Sepal Length", 
     col="green", 
     xlab="Sepal Length", 
     border="black")


```
```{r}

hist(iris$Petal.Length, 
     main="Histogram of petal Length", 
     col="green", 
     xlab="petal Length", 
     border="black")

```
```{r}

hist(iris$Sepal.Width, 
     main="Histogram of sepal width", 
     col="green", 
     xlab="sepal width", 
     border="black")

```
```{r}
hist(iris$Petal.Width, 
     main="Histogram of petal width", 
     col="green", 
     xlab="petal width", 
     border="black")

```
```{r}
# Density plot of Petal Length
# Density Plot (Smooth Distribution)
plot(density(iris$Petal.Length), 
     main="Density Plot of Petal Length", 
     col="green", 
     lwd=2)
     
```
```{r}
plot(density(iris$Petal.Width), 
     main="Density Plot of Petal Width", 
     col="green", 
     lwd=2)

```

```{r}
plot(density(iris$Sepal.Length), 
     main="Density Plot of Sepal Length", 
     col="green", 
     lwd=2)

```

```{r}
plot(density(iris$Sepal.Width), 
     main="Density Plot of Sepal Width", 
     col="green", 
     lwd=2)

```


```{r}
# Box Plot (Identifying Outliers)
# Boxplot of Sepal Width
boxplot(iris$Sepal.Width, 
        main="Box Plot of Sepal Width", 
        col="green")

```


```{r}
boxplot(iris$Petal.Width, 
        main="Box Plot of petalWidth", 
        col="green")
```

```{r}
boxplot(iris$Sepal.Length, 
        main="Box Plot of Sepal Length", 
        col="green")
```

```{r}
boxplot(iris$Petal.Length, 
        main="Box Plot of Petal length", 
        col="green")
```





# Exploring Categorical Data
```{r}
# Count of each species
table(iris$Species)

```

```{r}
# Bar plot for Species count
barplot(table(iris$Species), 
        main="Count of Each Species", 
        col=c("red", "green", "blue"))

```
# Multivariate Exploration
# Scatter Plot (Numeric vs. Numeric)
```{r}
# Scatter plot: Sepal Length vs. Sepal Width
plot(iris$Sepal.Length, iris$Sepal.Width, 
     main="Scatter Plot of Sepal Length vs Sepal Width",
     xlab="Sepal Length", ylab="Sepal Width",
     col=iris$Species, pch=19)
legend("topright", legend=levels(iris$Species), col=1:3, pch=19)

```

# Correlation (Numeric vs. Numeric)
```{r}
# Correlation between Sepal Length and Petal Length
cor(iris$Sepal.Length, iris$Petal.Length)

```
```{r}
cor(iris$Sepal.Width, iris$Petal.Width)
```


```{r}
# Correlation matrix
cor(iris[,1:4])

```

# Box Plot (Numeric vs. Categorical)
```{r}
# Box plot of Petal Length by Species
boxplot(Petal.Length ~ Species, data=iris,
        main="Box Plot of Petal Length by Species",
        col=c("red", "green", "blue"))

```
# Pair Plot (Multivariate Visualization)
```{r}
# Pair plot (Scatter plot matrix)
pairs(iris[,1:4], col=iris$Species, pch=19)
```
```{r}
boxplot(Sepal.Length ~ Species, data=iris,
        main="Box Plot of Sepal Length by Species",
        col=c("red", "green", "blue"))

```
```{r}
boxplot(Petal.Width ~ Species, data=iris,
        main="Box Plot of Petal  by Species",
        col=c("red", "green", "blue"))

```


```{r}
boxplot(Sepal.Width ~ Species, data=iris,
        main="Box Plot of Petal Length by Species",
        col=c("red", "green", "blue"))

```

```
# Using ggplot2 for Enhanced Visualization
```{r}
# Load ggplot2
library(ggplot2)
# Scatter plot with regression line
ggplot(iris, aes(x=Sepal.Length, y=Petal.Length, color=Species)) +
  geom_point() +
  geom_smooth(method="lm", se=FALSE) +
  ggtitle("Sepal Length vs Petal Length by Species")

```
